---
title: "Compartment Co-association with Density-Normalized Proximity: A Walkthrough"
author: "Dan Montagna"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: show
editor_options: 
  markdown: 
    wrap: 100
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Overview

This document explains, step-by-step, how the analysis pipeline builds per-cell “compartments” from
punctate objects (e.g., VPS35, EEA1), tests cross-type spatial association against a density-derived
null, summarizes per-well and per-genotype results, and returns both publication-grade plots and
exemplar cells for inspection. We will:

-   Load and wrangle raw spot files into the required schema.

-   Run ndrutils::compartment_coassoc() once (or across intensities).

-   Understand, in small chunks, how the function works:

    -   its geometry helpers,

    -   its density-normalized clustering (`DBSCAN`),

    -   its cross-type proximity test derived from a spatial Poisson model,

    -   its summaries, plots, and exemplars.

Throughout, we note assumptions, why the approach is reasonable, and why it stands up to scientific
rigor, with references to standard methods in spatial point pattern analysis and density-based
clustering.

# 1. End-user code

## 1.1 Libraries

```{r libraries}
library(tidyverse)
library(ndrutils)
library(readxl)
library(writexl)
library(cowplot)
library(ggforce)
```

**Conceptually** We use `tidyverse` for IO and wrangling, `cowplot`/`ggplot2` for visualization, and
`ndrutils` for the spatial analysis.

**Why this works**: `ndrutils::compartment_coassoc()` expects a tidy long data frame with specific
columns; `tidyverse` makes that shape easy to produce.

## 1.2 Parameters

We could do this down below, but I like to be able to set parameters right at the top of my markdown
file.

```{r parameters}

# ===================== PARAMETERS =============================================
px <- 0.149         # μm per pixel
ci.cutoff <- 10     # minimum corrected intensity for objects included
alpha_pair  <- 0.05 # significance for merging same-type puncta into compartments
alpha_cross <- 0.05 # significance for cross-type association (src -> tgt)

```

**Conceptually**

-   `px` lets us express distances in physical units later (μm).

-   `ci.cutoff` filters out low-confidence puncta.

-   `alpha_pair` controls how aggressively same-type puncta cluster into a “compartment”.

-   `alpha_cross` controls how stringent we are when calling a cross-type association significant.

**Why it works**: These alpha values are per-cell, density-normalized thresholds derived from a
homogeneous spatial Poisson process (details below). This makes decisions scale with the observed
density rather than fixed radii.

## 1.3 Load files and harmonize schema

You can start with data however you like, provided it has at least the following properties:

-   At least two types of objects that were masked in microscopy analysis. These are anonymously
    referred to as obj1 and obj2.

    -   In this example, I've masked puncta from immunostaining for either EEA1 or VPS35.

-   The following columns must also be present in the final dataframe:

    -   **well**: a character value that refers to the well (or slide / coverslip) from which the
        analyzed image was taken.
    -   **field**: a numeric value, useful for multiple images from a single well / slide /
        coverslip. If you have only one, it's ok to just indicate 1 - but this column is expected.
        -   *If this column does not exist in your data, just create it in the initial import pipe
            with `dplyr::mutate(field = as.numeric(1))`*
    -   **plane**: a numeric value, useful if you take multiple images in a z-stack. Same rules
        apply as for **field**
    -   **object.id**: whatever analysis software you used for puncta detection should have given
        each centroid a semi-unique identifier. In Revvity software like **Harmony** or **SimA**,
        these values are only unique *per image*. The code below combines this with the well name,
        field, and plane to create a value that uniquely identifies each object.
    -   **object.type** will be created with the code below. These are the names of obj1 and obj2.
    -   **x.coord**: X-coordinate for each object centroid.
    -   **y.coord** Y-coordinate for each object centroid.
    -   **corr.intensity**: The intensity of each puncta. This works best when corrected or
        normalized somehow during the image analysis itself, as different cutoffs cannot be set for
        the two types of objects. Future releases will be more flexible in this resepect.

```{r load data}
fpath <- "path\\to\\your\\data\\"

# Name of the subfolder holding .csv files for obj1
vps35.path <- paste0(fpath, "vps35")

# Name of the subfolder holding .csv files for obj1
eea1.path <- paste0(fpath, "eea1")


# Create a named list of lists.
# Each named list contains the names of all .csv files that describe the given object type
files_by_obj <- list(
  # obj1 is vps35 objects in this example
  vps35 = list.files(path = vps35.path, pattern = "*.csv", full.names = TRUE, ignore.case = TRUE),
  # obj2 is eea1 objects in this example
  eea1 = list.files(path = eea1.path, pattern = "*.csv", full.names = TRUE, ignore.case = TRUE)
)

# Helper function. For each of obj1 (vps35) and obj2 (eea1) read each .csv file into a single dataframe.
# Hold in a named list for the appropriate object.
read_and_bind <- function(file_vec) {
  purrr::map_dfr(file_vec, ~ readr::read_csv(.x, show_col_types = FALSE), .progress = TRUE)
}

# Run the helper function.
dfs_by_obj <- purrr::map(files_by_obj, read_and_bind, .progress = TRUE)


# The resulting dataframe has some really long column-names and a lot of unnecessary columns.
colnames(dfs_by_obj[[1]])


# Helper function coerces the messy dataframe into something we can work with.
# Once all the names are nice and tidy, we can row_bind() them together, since all columns will be named the same way.
make_df <- function(this.obj.name, this.df.in){
  dat <- this.df.in %>%
    dplyr::rename_with(~ tolower(.x)) %>%
    dplyr::rename(
      well           = wellname,
      cell           = 16,   # uses tidyselect position
      corr.intensity = 14,
      x.coord        = 20,
      y.coord        = 21
    ) %>%
    dplyr::mutate(
      unique.base = paste0(well, "_", plane, "_", field, "_"),   # per-object unique root
      unique.cell = paste0(unique.base, cell),
      object.id   = paste0(unique.base, object.number),
      object.type = this.obj.name,
      genotype    = dplyr::case_when(
        row %in% c(3, 4) ~ "WT",
        row %in% c(5, 6) ~ "KO",
        .default = NA_character_
      )
    ) %>%
    dplyr::select(genotype, well, field, unique.cell, object.id, object.type,
                  x.coord, y.coord, corr.intensity)

  return(dat)
}


# Form the tidy input dataframe

df <- purrr::imap(
  dfs_by_obj,
  ~ make_df(this.obj.name = .y, this.df.in = .x)
  ) %>%
  bind_rows

```

*Conceptually* We import two spot catalogs, harmonize column names, create unique identifiers for
cells and objects, assign object type and genotype, and select the columns required by
compartment_coassoc(): `genotype`, `well`, `field`, `unique.cell`, `object.id`, `object.type`,
`x.coord`, `y.coord`, and `corr.intensity.`

## 1.4 Run the analysis

```{r}

res <- ndrutils::compartment_coassoc(
  df,
  px = px,
  ci.cutoff = ci.cutoff,
  alpha_pair = alpha_pair,
  alpha_cross = alpha_cross
)

```

**Outputs**: When `length(ci.cutoff) == 1`, `res` is a list with:

```         
* `obj1_name`, `obj2_name`

* `well_summary` (long table of readouts by well/genotype)

* `final.plot` (2-panel cowplot: counts and percentages)

* `exemplars` (coordinates for selected cells close to genotype means)
```

You can draw the plot with: `res$final.plot`

# 2. Inside `compartment_coassoc()`, explained

**Below** we walk through the core function in small chunks.

**Idea**. We compute per-cell [compartments]{.underline} by clustering same-type puncta with a
**density-normalized DBSCAN radius**, then test [cross-type proximity]{.underline} against a Poisson
baseline to identify “significantly close” events, then summarize and visualize.

## 2.1 Input validation & type normalization

We ensure required columns exist and coerce key IDs to character so downstream `bind_rows()` won’t
fail mixing integers and strings (a common source of vctrs errors).

### Code (simplified)

```{r}

required_cols <- c("genotype","well","field","unique.cell","object.id",
                   "object.type","x.coord","y.coord","corr.intensity")
stopifnot(all(required_cols %in% names(df)))

df <- df %>%
  dplyr::mutate(
    genotype    = as.character(.data$genotype),
    well        = as.character(.data$well),
    field       = as.character(.data$field),
    unique.cell = as.character(.data$unique.cell)
  )

```

## 2.2 Geometry helpers: area & centroid of a cell

We estimate each cell’s footprint using the convex hull of all object coordinates in that cell: area
(px²) and centroid (px). Using a cell-specific area allows **density normalization per cel**l (λ =
N/Area). For degenerate cases (few points), we fall back to a minimal positive area and a simple
centroid to keep calculations stable.

### Code (simplified)

```{r}

.hull_area_px2 <- function(x, y) { ... }      # convex hull area
.hull_centroid_px <- function(x, y) { ... }   # polygon centroid

```

## 2.3 Key statistical link: the significance radius

### Code:

```{r}

.sig_radius <- function(lambda, alpha) {
  ifelse(lambda > 0, sqrt(-log1p(-alpha) / (pi * lambda)), Inf)
}


```

**Conceptually.** In a **2D homogeneous Poisson process** with intensity λ (points per px²), the
nearest-neighbor CDF is:

$$
\Pr(R \le r) = 1 - \exp(-\lambda \pi r^2).
$$

Solving for $r$ at a given tail probability $\alpha$ gives $$
r_\alpha = \sqrt{\frac{-\log(1-\alpha)}{\pi \lambda}},\quad 0<\alpha<1.
$$

**Why it works.** This yields a **density-normalized radius**: in denser fields, the same $\alpha$
produces a smaller $r_\alpha$. We use it twice:

-   **Pairing (same-type)**: as DBSCAN’s `eps` to merge nearby puncta into a compartment at
    significance level `alpha_pair`.

-   **Cross-type**: as the **threshold** below which a source compartment is called “significantly
    close” to its nearest target compartment, at level `alpha_cross`.

**Scientific rigor.** This is a well-known result in spatial point process theory (e.g., Illian et
al., 2008; Baddeley, Rubak & Turner, 2015). Using it aligns our calls with a model-based baseline
rather than arbitrary radii.

## 2.4 Build positive compartments (same-type clustering)

### Code (simplified)

```{r}

build_positive_compartments <- function(marker_df, cell_area_tbl, marker_name, alpha_pair, show_progress) {
  # per-cell loop
  # compute lambda = n / area
  r_eps <- .sig_radius(lambda, alpha_pair)
  # DBSCAN on (x,y) with eps = r_eps, minPts = 1
  cl <- dbscan::dbscan(coords, eps = r_eps, minPts = 1)
  # compute cluster centroids and sizes → one row per compartment
}

```

### Actual code:

```{r}


  build_positive_compartments <- function(marker_df, cell_area_tbl, marker_name,
                                          alpha_pair = 0.05, show_progress = TRUE) {
    # Coerce to data.table
    DT <- data.table::as.data.table(marker_df)
    CA <- data.table::as.data.table(cell_area_tbl)

    # Minimal column checks (fail fast with clear messages)
    req_DT <- c("genotype","well","field","unique.cell","x.coord","y.coord")
    miss_DT <- setdiff(req_DT, names(DT))
    if (length(miss_DT)) {
      stop("marker_df is missing required columns: ", paste(miss_DT, collapse = ", "))
    }
    req_CA <- c("unique.cell","area_px2")
    miss_CA <- setdiff(req_CA, names(CA))
    if (length(miss_CA)) {
      stop("cell_area_tbl is missing required columns: ", paste(miss_CA, collapse = ", "))
    }

    # Work per cell
    ucs <- unique(CA$unique.cell)
    out_list <- vector("list", length(ucs))

    if (isTRUE(show_progress)) {
      pb_id <- cli::cli_progress_bar(paste0("Building ", marker_name, " compartments"),
                                     total = length(ucs), .auto_close = FALSE, clear = FALSE)
      on.exit(try(cli::cli_progress_done(id = pb_id), silent = TRUE), add = TRUE)
    }

    for (i in seq_along(ucs)) {
      uc <- ucs[i]
      g  <- DT[unique.cell == uc]
      area_px2 <- CA[unique.cell == uc, area_px2][1]
      if (!is.finite(area_px2) || area_px2 <= 0) area_px2 <- 1e-6

      # No puncta for this cell -> return empty tibble with correct cols
      if (!nrow(g)) {
        out_list[[i]] <- tibble::tibble(
          genotype    = character(0),
          well        = character(0),
          field       = character(0),
          unique.cell = character(0),
          comp_id     = character(0),
          comp_x      = numeric(0),
          comp_y      = numeric(0),
          comp_size   = integer(0),
          marker      = character(0)
        )
        if (isTRUE(show_progress)) cli::cli_progress_update(id = pb_id, inc = 1)
        next
      }

      # Significance radius for same-type agglomeration
      lambda <- nrow(g) / area_px2
      r_eps  <- .sig_radius(lambda, alpha_pair)

      if (nrow(g) == 1L || !is.finite(r_eps) || r_eps <= 0) {
        # Single point => single-compartment centroid at that point
        cent <- tibble::tibble(
          genotype    = g$genotype[1],
          well        = g$well[1],
          field       = g$field[1],
          unique.cell = uc,
          comp_id     = paste0(uc, "__", marker_name, "__", 1L),
          comp_x      = g$x.coord[1],
          comp_y      = g$y.coord[1],
          comp_size   = 1L,
          marker      = marker_name
        )
        out_list[[i]] <- cent
        if (isTRUE(show_progress)) cli::cli_progress_update(id = pb_id, inc = 1)
        next
      }

      # ---- Defensive DBSCAN block --------------------------------------------
      coords <- as.matrix(g[, .(x.coord, y.coord)])
      if (nrow(coords) != nrow(g)) {
        stop("Internal error: coords rows (", nrow(coords), ") != g rows (", nrow(g), ") for cell ", uc)
      }

      cl <- dbscan::dbscan(coords, eps = r_eps, minPts = 1)

      cluster_vec <- as.integer(cl$cluster)
      if (length(cluster_vec) != nrow(g)) {
        cli::cli_alert_warning(
          "dbscan returned {length(cluster_vec)} clusters for {nrow(g)} points in cell {uc}; truncating to nrow(g)."
        )
        cluster_vec <- cluster_vec[seq_len(nrow(g))]
      }
      g[, cluster := cluster_vec]

      # Collapse points by cluster -> centroids
      cent <- g[, .(
        comp_x    = mean(x.coord),
        comp_y    = mean(y.coord),
        comp_size = .N
      ), by = .(genotype, well, field, unique.cell, cluster)][
        , `:=`(
          comp_id = paste0(unique.cell, "__", marker_name, "__", cluster),
          marker  = marker_name
        )][
          , .(genotype, well, field, unique.cell, comp_id, comp_x, comp_y, comp_size, marker)]

      out_list[[i]] <- tibble::as_tibble(cent)

      if (isTRUE(show_progress)) cli::cli_progress_update(id = pb_id, inc = 1)
    }

    dplyr::bind_rows(out_list)
  }
```

**Conceptually:**

-   Within each **cell** and **marker**, we compute λ and derive `r_eps`.

-   We run **DBSCAN** (Ester et al., 1996) with `eps = r_eps` and `minPts = 1`. Clusters represent
    **same-type compartments** that are **closer than expected by chance** under the Poisson null at
    level `alpha_pair`.

-   We then **average coordinates within each cluster** to record the **compartment centroid**
    (`comp_x`, `comp_y`) and **size**.

**Why it works:**

-   DBSCAN is **density-based** and does not require the number of clusters a priori. Using a
    **statistically grounded `eps`** aligns cluster formation with a **per-cell density scale**.

-   `minPts = 1` ensures solitary puncta are allowed to form a size-1 “compartment”, which is
    appropriate if density is low.

**Scientific rigor:**

-   DBSCAN is standard for discovering spatially contiguous clusters without convexity assumptions
    (Ester et al., 1996).

-   `eps` derived from the Poisson nearest-neighbor distribution ties cluster adjacency to a
    **formal null** rather than a heuristic distance.

-   Defensive coding handles corner cases (e.g., degenerate hulls, singletons, type coercion),
    improving reproducibility.

**References:**

-   Ester, M. et al. (1996) *A density-based algorithm for discovering clusters…* KDD.

-   Baddeley, A., Rubak, E., Turner, R. (2015) *Spatial Point Patterns: Methodology and Applications
    with R*. Chapman & Hall/CRC.

## 2.5 Within-type spacing diagnostics

### Code (simplified):

```{r}
add_within_nn_comp <- function(comp_tbl, cell_area_tbl) {
  # per cell+marker:
  # 1) compute each compartment’s nearest neighbor distance R
  # 2) compute rtilde = 2 * R * sqrt(lambda_comp)   (dimensionless)
}
```

### Actual code:

```{r}
  add_within_nn_comp <- function(comp_tbl, cell_area_tbl) {
    if (!nrow(comp_tbl)) {
      return(comp_tbl %>% dplyr::mutate(nn_comp = NA_real_, rtilde_comp = NA_real_))
    }
    keys <- c("genotype", "well", "field", "unique.cell", "marker")
    comp_tbl %>%
      dplyr::group_by(dplyr::across(dplyr::all_of(keys))) %>%
      dplyr::group_modify(~ {
        h <- .x
        uc <- .y$unique.cell[[1]]
        area_px2 <- cell_area_tbl$area_px2[match(uc, cell_area_tbl$unique.cell)]
        area_px2 <- ifelse(is.finite(area_px2) && area_px2 > 0, area_px2, 1e-6)

        if (nrow(h) < 2L) {
          h %>% dplyr::mutate(nn_comp = NA_real_, rtilde_comp = NA_real_)
        } else {
          nn <- RANN::nn2(
            data  = as.matrix(h[, c("comp_x", "comp_y")]),
            query = as.matrix(h[, c("comp_x", "comp_y")]),
            k = 2
          )
          R <- nn$nn.dists[, 2]
          lambda_comp <- nrow(h) / area_px2
          rtilde <- 2 * R * sqrt(lambda_comp)
          h %>% dplyr::mutate(nn_comp = R, rtilde_comp = rtilde)
        }
      }) %>%
      dplyr::ungroup()
  }
```

**Conceptually.** For QC/diagnostics, we compute **nearest-neighbor distances** among
**compartments** of the same marker within a cell, we report a density-scaled value
$\tilde r \propto R\sqrt{\lambda}$. This is a shape-free way to compare spacing across cells.

**Why it works.** Scaling by $\sqrt{\lambda}$ makes comparisons across densities more interpretable:
smaller $\tilde r$ indicates tighter-than-expected packing.

**Reference.** Nearest-neighbor functionals are standard tools (Illian et al., 2008; Baddeley et
al., 2015).

## 2.6 Cross-type association (directional, src → tgt)

### Code (simplified):

```{r}
classify_cross <- function(src_comp, tgt_comp, cell_area_tbl, alpha_cross) {
  # for each source compartment:
  #   Rst = distance to its nearest target compartment
  #   r_thr = .sig_radius(lambda_tgt, alpha_cross)
  #   is_close = (Rst <= r_thr)
}
```

### Code (actual):

```{r}

  classify_cross <- function(src_comp, tgt_comp, cell_area_tbl,
                             alpha_cross = 0.05, show_progress = TRUE) {
    if (!nrow(src_comp)) {
      return(src_comp %>%
               dplyr::mutate(nn_to_tgt = NA_real_, is_close = FALSE, r_thr_px = NA_real_))
    }
    ucs <- unique(src_comp$unique.cell)
    out <- vector("list", length(ucs))

    if (isTRUE(show_progress)) {
      pb_id <- cli::cli_progress_bar("Cross-type association", total = length(ucs),
                                     .auto_close = FALSE, clear = FALSE)
      on.exit(try(cli::cli_progress_done(id = pb_id), silent = TRUE), add = TRUE)
    }

    for (i in seq_along(ucs)) {
      uc <- ucs[i]
      s  <- src_comp %>% dplyr::filter(.data$unique.cell == uc)
      t  <- tgt_comp %>% dplyr::filter(.data$unique.cell == uc)
      area_px2 <- cell_area_tbl$area_px2[match(uc, cell_area_tbl$unique.cell)]

      if (!nrow(s) || !nrow(t)) {
        out[[i]] <- s %>%
          dplyr::mutate(nn_to_tgt = NA_real_, is_close = FALSE, r_thr_px = NA_real_)
        if (isTRUE(show_progress)) cli::cli_progress_update(id = pb_id, inc = 1)
        next
      }

      nn <- RANN::nn2(
        data  = as.matrix(t[, c("comp_x","comp_y")]),
        query = as.matrix(s[, c("comp_x","comp_y")]),
        k = 1
      )
      Rst <- as.numeric(nn$nn.dists[, 1])
      lambda_tgt <- nrow(t) / max(area_px2, 1e-6)
      r_thr <- .sig_radius(lambda_tgt, alpha_cross)   # pixels

      out[[i]] <- s %>%
        dplyr::mutate(
          nn_to_tgt = Rst,
          is_close  = is.finite(r_thr) & (Rst <= r_thr),
          r_thr_px  = r_thr
        )
      if (isTRUE(show_progress)) cli::cli_progress_update(id = pb_id, inc = 1)
    }

    dplyr::bind_rows(out)
  }
```

**Conceptually**: For each source compartment (e.g., VPS35) in a given cell, compute the distance
$R_{st}$ to its nearest target (e.g., EEA1) in that same cell. Estimate the target density $\lambda$
in that cell, compute $$
r_\alpha \;=\; \sqrt{\frac{-\ln(1-\alpha)}{\pi\,\lambda}},
$$ and declare the source–target pair **significantly close** if $R_{st} \le r_\alpha$.

**Why it works**: The test is directional: the proportion deemed close from source to target,
$\hat p_{\text{VPS35}\to\text{EEA1}}$, need not equal $\hat p_{\text{EEA1}\to\text{VPS35}}$ if the
per-cell densities differ. Using the **target** density in the threshold quantifies how “surprising”
a given proximity is given how many targets exist in that cell.

**Scientific rigor**: Under a homogeneous Poisson (PPP) null for target locations with intensity
$\lambda$ within a cell, the nearest-target distance from an arbitrary point has CDF $$
\Pr(R \le r)=1 - \exp(-\lambda\pi r^2).
$$ Thus $r_\alpha$ yields an exact size-$\alpha$ test at the per-source level: each source produces
a Bernoulli($\alpha$) decision under the null, and aggregating within a cell gives a
Binomial$(n_s,\alpha)$ model. This avoids global, cell-pooled radii and respects per-cell density
heterogeneity.

## 2.7 Summaries and labels

### Code (simplified):

```{r}

# per-cell counts:
n_type_comp    <- number_of_compartments_of_each_type
n_type_w_other <- number_of_source_compartments_called_close_to_target
p.type.w.other <- 100 * n_type_w_other / n_type_comp  # percent "close"

```

**Per-well aggregation**. We compute:

-   Counts per cell and convert to percent “close”.

-   Per-well mean ± SD of those cell-level values (used for bars and error bars).

-   Readable labels:

    -   p.<obj1>.w.<obj2> → “<obj1> compartments with <obj2> (signif.)”

    -   n.<obj> → “<obj>+” (average number of compartments)

**Why it works**: Per-well aggregation matches common experimental designs; mean ± SD communicates
central tendency and dispersion across fields/cells.

**Note on statistics**: The plotted error bars are summary SDs, not confidence intervals. For
inference across genotypes/wells, use appropriate models (e.g., mixed models, permutation tests).

## 2.8 Visualization

### Code

```{r}

    # plotting
    bargraphing <- function(type_code) {
      p <- well_summary %>%
        dplyr::filter(.data$readout.type %in% type_code) %>%
        ggplot2::ggplot(ggplot2::aes(x = .data$genotype, y = .data$percent, fill = .data$genotype)) +
        ggplot2::facet_wrap(~ readout, nrow = 1, scales = "free_y") +
        ggplot2::stat_summary(fun = mean, geom = "col", width = 0.6, color = "white", alpha = 0.65) +
        ggplot2::stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.18, linewidth = 0.6) +
        ggplot2::geom_point(
          position = ggplot2::position_jitter(width = 0.10, height = 0, seed = 1), size = 1
        ) +
        theme_ndrutils(base_size = 8) +
        ggplot2::theme(
          plot.title = ggplot2::element_text(hjust = 0.5, margin = ggplot2::margin(b = 5)),
          axis.text.y = ggplot2::element_text(hjust = 1.5)
        )

      if (type_code == "p") {
        p <- p +
          ggplot2::scale_y_continuous(
            labels = function(x) paste0(round(x, 1), "%"),
            expand = ggplot2::expansion(mult = c(0, 0.1))
          ) +
          ggplot2::labs(
            title = "Compartment co-association (significantly closer-than-random)",
            x = NULL, y = "% of compartments", fill = "Genotype", color = "Well"
          )
      } else if (type_code == "n") {
        p <- p +
          ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0, 0.1))) +
          ggplot2::labs(
            title = "Number of compartments (single identity)",
            x = NULL, y = "# of compartments", fill = "Genotype", color = "Well"
          )
      }

      genos <- well_summary %>% dplyr::distinct(.data$genotype) %>% dplyr::pull()
      genos <- as.character(genos[!is.na(genos)])
      if (length(genos) == 2L && requireNamespace("ggsignif", quietly = TRUE)) {
        p <- p +
          ggsignif::geom_signif(
            stat = "signif",
            comparisons = list(genos),
            map_signif_level = TRUE,
            textsize = 4, margin_top = 0.3, vjust = 0.2
          )
      } else {
        if (length(genos) > 2L) {
          cli::cli_alert_info("More than two genotypes detected; omitting significance brackets.")
        } else if (!requireNamespace("ggsignif", quietly = TRUE)) {
          cli::cli_alert_info("Package 'ggsignif' not available; omitting significance brackets.")
        }
      }
      p
    }

    data_plots <- purrr::map(c("n", "p"), bargraphing)

    titleplot <- cowplot::ggdraw() +
      cowplot::draw_label(
        paste0("Corrected Intensity = ", ci, " AU"),
        fontface = "bold", hjust = 0.5
      )

    plotlist <- c(list(titleplot), data_plots)

    final.plot <- cowplot::plot_grid(
      plotlist = plotlist,
      ncol = 1,
      rel_heights = c(title_rel_height, rep(1, length(plotlist) - 1))
    )


```

The two-row panel shows:

-   **Top**: number of compartments per object (workload/abundance proxy).

-   **Bottom**: % of source compartments with significant proximity to target (main readout).

The paired view guards against misinterpretation: a high percentage with very few compartments may
be fragile; conversely, many compartments with a moderate percentage can still be biologically
meaningful.

## 2.9 Exemplars (For QC & communication)

### Code:

```{r}
# For each readout × genotype:
# 1) compute genotype mean at the cell level
# 2) pick up to 5 cells closest to that mean (small absolute deviation)
# 3) export centered coordinates (μm) for obj1/obj2/comp1/comp2

    # --------- EXEMPLARS ----------
    exemplars <- NULL
    if (isTRUE(return_exemplars)) {
      cell_long <- cell_summary %>%
        tidyr::pivot_longer(
          cols = dplyr::all_of(c(p1_name, p2_name, n1_name, n2_name)),
          names_to = "readout_name", values_to = "value"
        ) %>%
        dplyr::mutate(
          readout_label = dplyr::case_when(
            .data$readout_name == p1_name ~ p1_label,
            .data$readout_name == p2_name ~ p2_label,
            .data$readout_name == n1_name ~ n1_label,
            .data$readout_name == n2_name ~ n2_label,
            TRUE ~ .data$readout_name
          )
        )

      geno_means <- cell_long %>%
        dplyr::group_by(.data$genotype, .data$readout_name, .data$readout_label) %>%
        dplyr::summarise(mean_value = mean(.data$value, na.rm = TRUE), .groups = "drop")

      closest_cells <- cell_long %>%
        dplyr::left_join(geno_means, by = c("genotype","readout_name","readout_label")) %>%
        dplyr::filter(is.finite(.data$value), is.finite(.data$mean_value)) %>%
        dplyr::mutate(dev = abs(.data$value - .data$mean_value)) %>%
        dplyr::arrange(.data$genotype, .data$readout_label, .data$dev, .data$unique.cell) %>%
        dplyr::group_by(.data$genotype, .data$readout_label) %>%
        dplyr::slice_head(n = 5) %>%
        dplyr::ungroup()

      centers_idx <- cell_geom %>%
        dplyr::select(.data$unique.cell, .data$center_x_px, .data$center_y_px)

      build_cell_exemplar <- function(uc) {
        cx <- centers_idx$center_x_px[match(uc, centers_idx$unique.cell)]
        cy <- centers_idx$center_y_px[match(uc, centers_idx$unique.cell)]
        if (!is.finite(cx)) cx <- 0
        if (!is.finite(cy)) cy <- 0

        obj1 <- obj1_df %>%
          dplyr::filter(.data$unique.cell == uc) %>%
          dplyr::transmute(
            genotype, well, field, unique.cell,
            kind = "obj1",
            id   = .data$object.id,
            x_rel_um = (.data$x.coord - cx) * px,
            y_rel_um = (.data$y.coord - cy) * px,
            comp_size = as.integer(NA),
            is_close  = NA,
            r_thr_px  = NA_real_
          )

        obj2 <- obj2_df %>%
          dplyr::filter(.data$unique.cell == uc) %>%
          dplyr::transmute(
            genotype, well, field, unique.cell,
            kind = "obj2",
            id   = .data$object.id,
            x_rel_um = (.data$x.coord - cx) * px,
            y_rel_um = (.data$y.coord - cy) * px,
            comp_size = as.integer(NA),
            is_close  = NA,
            r_thr_px  = NA_real_
          )

        c1 <- comp1_cls %>%
          dplyr::filter(.data$unique.cell == uc) %>%
          dplyr::transmute(
            genotype, well, field, unique.cell,
            kind = "comp1",
            id   = .data$comp_id,
            x_rel_um = (.data$comp_x - cx) * px,
            y_rel_um = (.data$comp_y - cy) * px,
            comp_size = as.integer(.data$comp_size),
            is_close  = .data$is_close,
            r_thr_px  = .data$r_thr_px
          )

        c2 <- comp2_cls %>%
          dplyr::filter(.data$unique.cell == uc) %>%
          dplyr::transmute(
            genotype, well, field, unique.cell,
            kind = "comp2",
            id   = .data$comp_id,
            x_rel_um = (.data$comp_x - cx) * px,
            y_rel_um = (.data$comp_y - cy) * px,
            comp_size = as.integer(.data$comp_size),
            is_close  = .data$is_close,
            r_thr_px  = .data$r_thr_px
          )

        dplyr::bind_rows(obj1, obj2, c1, c2)
      }

      exemplars <- closest_cells %>%
        dplyr::group_by(.data$readout_label, .data$genotype) %>%
        dplyr::summarise(
          cells = list(unique(.data$unique.cell)),
          .groups = "drop"
        ) %>%
        purrr::pmap(function(readout_label, genotype, cells) {
          dfs <- purrr::map(cells, build_cell_exemplar)
          list(readout = readout_label, genotype = genotype, dfs = dfs)
        }) %>%
        {
          out <- list()
          for (elt in .) {
            if (is.null(out[[elt$readout]])) out[[elt$readout]] <- list()
            names(elt$dfs) <- as.character(seq_along(elt$dfs))
            out[[elt$readout]][[as.character(elt$genotype)]] <- elt$dfs
          }
          out
        }
    }

```

**Conceptually**: We present representative cells whose readout is closest to the genotype mean,
with centered coordinates (μm) for visual overlays. This aids communication/QC, not inference.

**Why it works**: Selecting near the mean avoids cherry-picked extremes and helps readers see
“typical” examples.

## 2.10 Multiple intensity cutoffs

If `ci.cutoff` is a vector, the pipeline is run for each cutoff and returns a named list (e.g.,
`ci=0`, `ci=10`), carrying object names as attributes for consistent labeling.

## 3. Scientific assumptions, robustness, and alternatives

### 3.1 Poisson baseline and density normalization

-   We assume a homogeneous spatial Poisson process within each cell as a null for both same-type
    agglomeration (`alpha_pair`) and cross-type proximity (`alpha_cross`).

-   This yields the closed-form nearest-neighbor distribution and the $\alpha$-level radii used
    above.

-   

**Why reasonable.** As a null, a homogeneous Poisson is a standard choice for spatial randomness
absent interaction (Illian et al., 2008; Baddeley et al., 2015). Using **cell-specific** λ adapts to
heterogeneity in object abundance and cell size.

**Caveats.**

-   Strong inhomogeneities (e.g., gradients) suggest an inhomogeneous PPP; see alternatives below.

-   If points interact (repulsion/attraction), second-order functionals (Ripley’s $K$, cross-$K$) or
    permutation tests can complement this analysis.

### 3.2 Why DBSCAN with Poisson-derived $\epsilon$

-   DBSCAN finds contiguous groups without assuming cluster shape and is robust to noise (Ester et
    al., 1996).

-   Setting $\epsilon$ by the Poisson $\alpha$-radius calibrates “what counts as close” to the
    observed density, avoiding arbitrary pixel thresholds.

### 3.3 Why convex hull for area

-   The convex hull provides a conservative estimate of spatial support using only point
    coordinates.

-   If cell segmentations are available, substituting segmented areas for $\lambda$ can reduce bias
    when points sample only part of the cell.

### 3.4 Alternatives and extensions

-   **Permutation within cell:** permute type labels to form an empirical null for
    cross-association, preserving each type’s spatial autocorrelation.

-   **Inhomogeneous PPP:** model spatial trend in intensity and re-derive thresholds.

-   **Cross-**$K$ pair correlation: second-order analyses for interactions across scales.

-   **Mixed-effects models:** genotype comparisons accounting for plate/field structure.

## 4. Practical tips

-   Keep units clear (px vs μm); store `px` in metadata when saving exemplars.

-   Inspect edge cases: cells with very few objects (singletons allowed) and extreme densities
    (check $r_\varepsilon$).

-   Validate genotype mapping and well/field joins early to avoid label leakage.

-   Run sensitivity analyses on `ci.cutoff`, `alpha_pair`, and `alpha_cross`.
